Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-25 paper.tex
Responses to comments

The properties of the inferred ARGs are summarised in Table~\ref{tab:args}.
% TODO more discussion here, making sure we explain anything that might not be
% obvious.
Most of the nodes in the ARGs represent sample genomes
(Wide ARG: 87\%, Long ARG: 84\%), but the inferred ancestral sequences
represented by the nodes inferred from daily sample clusters
by UPGMA (Section~\ref{sec:sample-cluster-tree-inference})
and parsimony heuristics (Section~\ref{sec:parsimony-heuristics})
contribute significantly.
Table~\ref{tab:args} shows the counts of edges (see
Section~\ref{sec:args} for a technical definition) in the ARGs
categorised by parent node type, and demonstrates that each
of these contributes significantly to the overall topology.
In particular, we can see that approximately 40\% of the edges
in both ARGs involve inheriting directly from sampled genomes.
Both ARGs contain 29,422 sites (Section \ref{sec:data_preprocessing}),
and a large number of mutations.
The average number of mutations per-site is high, although much
of this may be explained by outlier sites with artefactually high
mutation counts (see Figure~\ref{fig:breakpoint-distribution}).
Despite this, however, the number of mutations per node
is small. In the Wide ARG, for example, we have a mean of 0.77
mutations per sample, demonstrating that new samples are generally
well represented by it. See Section~\ref{sec:mutation_spectrum}
for more analysis of the patterns of inferred mutations.
Recombination plays a relatively minor role, with $<0.3\%$
of nodes in the ARGs representing inferred recombination events.
Of these recombination nodes, the majority are ancestral
to only one sample (Wide ARG: FIXME\%, Long ARG: FIXME \%).
We analyse signals of recombination in Sections~\ref{sec:jackson_recombs},
\ref{sec:breakpoint_intervals}, \ref{sec:parent_divergence},
and \ref{sec:pango_x_lineages}.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/wide_arg_recombination_intervals.pdf}
\caption{\label{fig:breakpoint-distribution}
Distribution of recombination breakpoints and mutations along the genome in
the Wide ARG. Top panel shows the intervals for 1,769 breakpoints associated
with 1,522 recombination nodes with at least two descending samples, plotted along the genome
as line segments (coloured by interval width). The inset histogram shows the
distribution of these interval widths.
The bottom panel shows the number of intervals that intersect with
each site along the genome (left axis, orange), and the number of mutations
per site (right axis, blue).
The top-ten sites by mutation count are annotated.
See Figure~\ref{fig:long_arg_breakpoint_distribution} for the equivalent plot
for the Long ARG.}
\end{figure}

@@

\section{Discussion}
% PARA 1. SARS-CoV-2 is still important, and the vast hordes of data provide an
% unprecedented opportunity to study the evolution of a virus. Even if the
% public health emergency is officially over, it doesn't mean it's not coming
% back and we should be prepared. Recombination is now a vital factor, and
% present a way in which it can be incorporated. Our results are accurate.

The COVID-19 pandemic continues to be a global emergency [DOES IT? Didn't the
WHO downgrade recently?]
with persistent high levels of infection. This high prevalence has allowed the
proliferation of many variants, with more than 600 Pango-designated lineages
circulating globally in the last three months (January to March, 2023;
https://gisaid.org/; access on March 27, 2023). High prevalence also brings a
higher risk of coinfection, increasing opportunities for new phenotypically
distinct recombinants to emerge and spread. Phylogenetic approaches have
been central to pandemic responses thus far [TODO: Examples and citations].
However, as recombinant lineages are now endemic and it is clear that
recombination is a major feature of ongoing SARS-CoV-2 evolution,
it is imperative that these methods are updated to incorporate the
effects of recombination, and future public health interventions are
not based on incomplete and potentially biased evolutionary models.
Here we have introduced the first method to infer an evolutionary
history that jointly models genetic inheritance, mutation and
recombination at pandemic scale, and illustrated how this single
structure accurately captures results derived by many different means.

>>>>

The properties of the inferred ARGs are summarised in Table~\ref{tab:args}.
% TODO more discussion here, making sure we explain anything that might not be
% obvious.
The majority of the nodes in the ARGs represent sample genomes
(Wide ARG: 87\%, Long ARG: 84\%), with the remainder mostly representing
the ancestral sequences inferred by UPGMA 
from daily sample clusters (Section~\ref{sec:sample-cluster-tree-inference})
and parsimony heuristics (Section~\ref{sec:parsimony-heuristics}).
Table~\ref{tab:args} shows the counts of edges (see
Section~\ref{sec:args} for a technical definition) in the ARGs
categorised by parent node type, and demonstrates that each
of these contributes significantly to the overall topology.
In particular, we can see that approximately 40\% of the edges
in both ARGs involve inheriting directly from sampled genomes.
Both ARGs contain 29,422 sites (Section \ref{sec:data_preprocessing}),
and a large number of mutations.
The average number of mutations per-site is high, although much
of this may be explained by outlier sites with artefactually high
mutation counts (see Figure~\ref{fig:breakpoint-distribution}).
Despite this, however, the number of mutations per node
is small. In the Wide ARG, for example, we have a mean of 0.77
mutations per sample, demonstrating that new samples are generally
well represented by it. See Section~\ref{sec:mutation_spectrum}
for more analysis of the patterns of inferred mutations.
Recombination plays a relatively minor role, with $<0.3\%$
of nodes in the ARGs representing inferred recombination events.
Of these recombination nodes, the majority are ancestral
to only one sample (Wide ARG: FIXME\%, Long ARG: FIXME \%).
We analyse signals of recombination in Sections~\ref{sec:jackson_recombs},
\ref{sec:breakpoint_intervals}, \ref{sec:parent_divergence},
and \ref{sec:pango_x_lineages}.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/wide_arg_recombination_intervals.pdf}
\caption{\label{fig:breakpoint-distribution}
Distribution of recombination breakpoints and mutations along the genome in
the Wide ARG. Top panel shows the intervals for 1,769 breakpoints associated
with 1,522 recombination nodes with at least two descending samples, plotted along the genome
as line segments (coloured by interval width). The inset histogram shows the
distribution of these interval widths (truncated at 10kb).
The bottom panel shows the number of intervals that intersect with
each site along the genome (left axis, orange), and the number of mutations
per site (right axis, blue).
The top-ten sites by mutation count are annotated.
See Figure~\ref{fig:long_arg_breakpoint_distribution} for the equivalent plot
for the Long ARG.}
\end{figure}

@@

\section{Discussion}
% PARA 1. SARS-CoV-2 is still important, and the vast hordes of data provide an
% unprecedented opportunity to study the evolution of a virus. Even if the
% public health emergency is officially over, it doesn't mean it's not coming
% back and we should be prepared. Recombination is now a vital factor, and
% present a way in which it can be incorporated. Our results are accurate.
Although the COVID-19 pandemic is no longer considered a global emergency
by the WHO, the prevalence of SARS-CoV-2 continues to be high worldwide
with persistent high levels of infection. This high prevalence has allowed the
proliferation of many variants, with more than 600 Pango-designated lineages
circulating globally in the last three months (January to March, 2023;
https://gisaid.org/; access on March 27, 2023). High prevalence also brings a
higher risk of coinfection, increasing opportunities for new phenotypically
distinct recombinants to emerge and spread. Phylogenetic approaches have
been central to pandemic responses thus far [TODO: Examples and citations].
However, as recombinant lineages are now endemic and it is clear that
recombination is a major feature of ongoing SARS-CoV-2 evolution,
it is imperative that these methods are updated to incorporate the
effects of recombination, and future public health interventions are
not based on incomplete and potentially biased evolutionary models.
Here we have introduced the first method to infer an evolutionary
history that jointly models genetic inheritance, mutation and
recombination at pandemic scale, and illustrated how this single
structure accurately captures results derived by many different means.
